---
title: "W#03 Amounts Proportions Uncertainty"
subtitle: "With inspiration from the Data Visualization Courses of Andrew Heiss and Data Science Box from Mine Cetinkaya-Rundel"
author: Jan Lorenz
format: 
  revealjs: 
    slide-number: true
    preview-links: false
    logo: img/ConstructorUniversity.png
    footer: "MDSSB: Visual Communication and Data Storytelling"
    smaller: true
    code-fold: true
    echo: true
---

# Visualize Amounts and Proportions by Example

## Brexit Opinions in the UK

History:

- In 2016, the UK voted to leave the EU in a [referendum](https://en.wikipedia.org/wiki/2016_United_Kingdom_European_Union_membership_referendum) with 51.8% for leaving and 48.1% for remaining.
- In 2020, the UK formally left the EU after long negotiations about the exact condition. 
- Inbetween, there were elections, changes in government and a heated debate.

- In 2019, a YouGov survey asked people whether they think leaving the EU was the right or wrong decision.

Let us work with this data to illustrate principles of effective visualizations.

## Brexit Survey Data

In September 2019, YouGov survey asked 1,639 adults in the UK the following question:

:::{style="font-size:80%;"} 
*In hindsight, do you think Britain was right/wrong to vote to leave EU?*   
🔲 Right to leave | 🔲 Wrong to leave | 🔲 Don't know
:::

The file `brexit_opinions.csv` in the  dataset in the repository [VisCom_Playground](https://github.com/CU-F25-MDSSB-MET-03-VisCommDataStory/VisCom_Playground). 

::: {.columns}
::: {.column}
Look at the first 10 rows
```{r}
library(tidyverse)
brexit <- read_csv("../data/brexit_opinions.csv")
brexit
```

:::

::: {.column .fragment}
Look at 10 random rows

```{r}
brexit |> slice_sample(n = 10)
```

:::
:::

## First basic visuals with `geom_bar`

::: {.columns}
::: {.column}

```{r}
ggplot(brexit, aes(x = opinion)) +
  geom_bar()
```

:::
::: {.column}

```{r}
ggplot(brexit, aes(x = region)) +
  geom_bar()
```

:::
:::

## `geom_bar` vs. `geom_col`

- `geom_bar()` counts the number of occurrences of each category in the data and creates bars based on these counts.
    - It is a geom which also does some data transformation! (Here counting)
- `geom_col()` requires you to provide the heights of the bars explicitly. It uses the values in the data to determine the height of each bar.
    - It is the basic building block for bar plots


## `geom_bar` vs. `geom_col`

::: {.columns}
::: {.column width="33%"}
With `geom_bar`

```{r}
brexit |>
  ggplot(aes(x = opinion)) +
  geom_bar()
```

:::
::: {.column width="33%"}
For `geom_col` we need to count first

```{r}
brexit |>
  count(opinion)
```

:::
::: {.column width="33%"}
With `geom_col` we must specify `x` and `y`

```{r}
brexit |>
  count(opinion) |>
  ggplot(aes(x = opinion, y = n)) +
  geom_col()
```

:::
:::

. . . 


Alphabetical order is rarely what you want! Do we really want "Don't know" first?



## Order Matters

Often good: Order by frequency

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(x = fct_infreq(opinion))) +
  geom_bar()
```

. . . 

`fct_infreq`: (Creates and) reorders a factor by the frequency. 


## Clean up labels

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(x = fct_infreq(opinion))) +
  geom_bar() +
  labs(
    x = "Opinion",
    y = "Count"
  )
```


## Order Matters: Inherent levels 

Reorder **manually** with a custom order

`fct_relevel`: Reorders the levels of a factor as specified

```{r}
#| echo: true
brexit <- brexit |>
  mutate(
    region = fct_relevel(
      region,
      "london",
      "rest_of_south",
      "midlands_wales",
      "north",
      "scot"
    )
  )
```

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(x = region)) +
  geom_bar()
```



## Clean factor labels 

Reorder manually with a custom order

`fct_recode`: Change factor levels by hand

```{r}
#| echo: true
brexit <- brexit |>
  mutate(
    region = fct_recode(
      region,
      London = "london",
      `Rest of South` = "rest_of_south",
      `Midlands / Wales` = "midlands_wales",
      North = "north",
      Scotland = "scot"
    )
  )
```

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(x = region)) +
  geom_bar()
```



## Long categories names on the y-axis!


```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(y = region)) +
  geom_bar()
```


## Reverse the order of levels to start with the top

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(y = fct_rev(region))) +
  geom_bar()
```



## Clean up axis lables 

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(y = fct_rev(region))) +
  geom_bar() +
  labs(
    x = "Count",
    y = ""
  )
```

Do we really need "Region" on the y-axis?



## Pick a purpose!

What is the (main) purpose of the visualization?  

. . . 

Purpose: **Show how support for Brexit differs regionally**


## First try: Segmented stacked barplot

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = region)) +
  geom_bar()
```

## Or "the other way round"?

```{r}
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
ggplot(brexit, aes(y = region, fill = opinion)) +
  geom_bar()
```

. . . 

They are both hard to read.


## Use facets

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = region)) +
  geom_bar() +
  facet_wrap("region", nrow = 1)
```


## Avoid redundancy?

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1)
```

## Redundancy can help to tell a story. 

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1)
```

## Be selective with redundancy

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none")
```

We do not need a legend when the colors are also `y`-labels.


## Use informative labels

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none") +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    x = NULL,
    y = NULL
  )
```


## More information (Survey and Source)

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none") +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL
  )
```



## Select meaningful colors

Are the default rainbow colors meaningful?
Another option: Pick colors manually

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none") +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "red",
      "Right" = "green",
      "Don't know" = "gray"
    )
  )
```



## Choose better colors

[colorbrewer2.org](https://colorbrewer2.org){preview-link="true"}


## Use better colors

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none") +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  )
```



## Choose a theme

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none") +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  theme_minimal()
```


## Adjust the general text size

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap("region", nrow = 1) +
  guides(fill = "none") +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  theme_minimal(base_size = 14)
```


## General principles	applied before

- Order matters
- Put long categories on the y-axis
- Use meaningful labels
- **Pick a purpose**
- Select meaningful colors


## What about percentages?

Let us first compute percentages. But what percentages?

::: {.columns}
::: {.column}

```{r}
#| code-fold: false
brexit |>
  count(opinion, region)
```

:::
::: {.column}

```{r}
#| code-fold: false
brexit |>
  count(region, opinion)
```

:::
::: 

. . . 

Sensible would be percentages within each region.

## Compute percentages within regions

```{r}
#| code-fold: false

brexit |>
  count(region, opinion) |>
  mutate(pct = n / sum(n) * 100, .by = region)
```

## Stacked bar plot with percentages

```{r}
#| fig-height: 4
#| fig-width: 8

brexit |>
  count(region, opinion) |>
  mutate(pct = n / sum(n) * 100, .by = region) |>
  ggplot(aes(y = fct_rev(region), x = pct, fill = opinion)) +
  geom_col() +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = "Percentage",
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  theme_minimal(base_size = 14)
```

Is the order of Wrong/Right/Don't know good here?

## Stacked bar plot with percentages

```{r}
#| fig-height: 4
#| fig-width: 8

brexit |>
  count(region, opinion) |>
  mutate(opinion = fct_relevel(opinion, "Right", "Don't know", "Wrong")) |>
  mutate(pct = n / sum(n) * 100, .by = region) |>
  ggplot(aes(y = fct_rev(region), x = pct, fill = opinion)) +
  geom_col() +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = "Percentage",
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 14)
```


## Elegant percentages with `scales`

Also make bars a bit thinner.

```{r}
#| fig-height: 4
#| fig-width: 8

library(scales)
brexit |>
  count(region, opinion) |>
  mutate(opinion = fct_relevel(opinion, "Right", "Don't know", "Wrong")) |>
  mutate(pct = n / sum(n), .by = region) |>
  ggplot(aes(y = fct_rev(region), x = pct, fill = opinion)) +
  geom_col(width = 0.75) +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = "",
    y = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 14)
```

## Bar plots vs. Pie Charts

::: {.columns}
::: {.column}

```{r}
#| fig-height: 5
brexit |>
  count(opinion) |>
  ggplot(aes(x = "", y = n, fill = fct_rev(opinion))) +
  geom_col() +
  coord_polar(theta = "y") +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL,
    fill = "Opinion"
  ) +
  theme_void(base_size = 16)
```

:::
::: {.column}

```{r}
#| fig-height: 5
brexit |>
  count(opinion) |>
  ggplot(aes(x = fct_rev(opinion), y = n, fill = fct_rev(opinion))) +
  geom_col() +
  scale_fill_manual(
    values = c(
      "Wrong" = "#ef8a62",
      "Right" = "#67a9cf",
      "Don't know" = "gray"
    )
  ) +
  labs(
    title = "Was Britain right/wrong to vote to leave EU?",
    subtitle = "YouGov Survey Results, 2-3 September 2019",
    x = NULL,
    y = NULL,
    fill = "Opinion"
  ) +
  theme_minimal(base_size = 16)
```

:::
:::

What is easier to read for what purpose?

## \#barbarplots

{{< video https://www.youtube.com/watch?v=LFDbqw2xPbQ  width="600" height="400" >}}

::: aside
<https://barbarplots.github.io/index.html>
:::


# Uncertainty and Distributions



## Visualize all data before aggreagteing 

Lets test the dogs and cats example with self created data. 

```{r}
set.seed(1234)
animals <- tibble(
  animal = c(rep(c("Small cat", "Big cat"), each = 250), rep("Dog", 500))
) |>
  mutate(
    weight = case_when(
      animal == "Small cat" ~ rnorm(n(), 20, 5),
      animal == "Big cat" ~ rnorm(n(), 60, 5),
      animal == "Dog" ~ rnorm(n(), 40, 10)
    )
  ) |>
  mutate(animal_type = ifelse(str_detect(animal, "cat"), "Cats", "Dogs"))

animals_mean <- animals |>
  group_by(animal_type) |>
  summarize(avg_weight = mean(weight))

ggplot(animals_mean, aes(x = animal_type, y = avg_weight, fill = animal_type)) +
  geom_col() +
  labs(x = NULL, y = "Weight") +
  guides(fill = "none")
```


## Show all data with jitter plots

```{r}
ggplot(animals, aes(x = animal_type, y = weight, color = animal_type)) +
  geom_jitter() +
  geom_point(
    data = animals_mean,
    aes(y = avg_weight),
    color = "black",
    size = 5
  ) +
  labs(x = NULL, y = "Weight") +
  guides(color = "none")
```

## Add the averages and make points transparent

```{r}
ggplot(animals, aes(x = animal_type, y = weight, color = animal_type)) +
  geom_jitter(alpha = 0.5) +
  geom_point(data = animals_mean, aes(y = avg_weight), size = 10) +
  labs(x = NULL, y = "Weight") +
  guides(color = "none")
```

Note: We include a different dataset for the averages with `data = animals_mean` in `geom_point()`

## Show all data with beeswarm plots

```{r}
library(ggbeeswarm)

ggplot(animals, aes(x = animal_type, y = weight, color = animal_type)) +
  geom_beeswarm(size = 1) +
  # Or try this too:
  # geom_quasirandom() +
  labs(x = NULL, y = "Weight") +
  guides(color = "none")
```

## Combine boxplots with points

```{r}
ggplot(animals, aes(x = animal_type, y = weight, color = animal_type)) +
  geom_boxplot(width = 0.5) +
  geom_point(position = position_jitter(height = 0), size = 1, alpha = 0.5) +
  labs(x = NULL, y = "Weight") +
  guides(color = "none")
```

## Combine violins with points

```{r}
ggplot(animals, aes(x = animal_type, y = weight, color = animal_type)) +
  geom_violin(width = 0.5) +
  geom_point(position = position_jitter(height = 0), size = 1, alpha = 0.5) +
  labs(x = NULL, y = "Weight") +
  guides(color = "none")
```


## Overlapping ridgeplots

```{r}
library(ggridges)

ggplot(animals, aes(x = weight, y = animal_type, fill = animal_type)) +
  geom_density_ridges() +
  labs(x = "Weight", y = NULL) +
  guides(fill = "none")
```

## More information is better

- Avoid ploting single numbers when there is a whole range behind it  
- OK, you may do it in a longer data story after making clear how the variation looks like



# How to select a visual? 

"Flow Chart" Website with examples and links to code in `R` and `python`:

[data-to-viz.com](https://www.data-to-viz.com/)



## Practice: Distribution Visualization

Do some of visualizations of distributions using the country dataset in the [VisCom_Playground](https://github.com/CU-F25-MDSSB-MET-03-VisCommDataStory/VisCom_Playground) repository.

Create a new slidedeck with some trials. 

For example: Use Life Expectancy and compare the distribution of countries' value between continents. You can also visualize how a distribution changes over time.